---
title: "Reproducible Research: Peer Assessment 1"
author: "Fengrong Yang"
date: "12/14/2020"
output: 
  html_document:
    keep_md: true
---
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](https://www.fitbit.com/global/us/home/), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband/), or [Jawbone Up](https://jawbone.com/up/). These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:  
* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)[52K]  

The variables included in this dataset are:  
* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as \color{red}{\verb|NA|}NA)  
* **date**: The date on which the measurement was taken in YYYY-MM-DD format  
* **interval**: Identifier for the 5-minute interval in which measurement was taken  
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.  

## Loading and preprocessing the data
### 1. Unzip the data to obtain a csv file.
```{r unzip, cache = TRUE}
unzip("activity.zip", exdir = "data")
```
### 2. Read the data into activity data frame and show summary statistics
```{r reading, cache = TRUE}
activity <- read.csv("data/activity.csv", stringsAsFactors = FALSE)
str(activity)
summary(activity)
```
### 3. Convert date to POSIXct class using lubridate package and convert interval to hour:minute format
```{r date_conversion, message = FALSE}
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
```

## What is mean total number of steps taken per day?
### 1. Calculate the total number of steps taken per day
```{r total_per_day, message = FALSE}
require(dplyr)
total_per_day <- activity %>% group_by(date) %>% summarise(total_steps = sum(steps, na.rm = TRUE), na = mean(is.na(steps))) %>% print
```
Visualize the total number of steps taken per day as a barplot:
```{r barplot, fig.width = 12}
barplot(height = total_per_day$total_steps, names.arg = total_per_day$date, cex.names = 0.68, las = 3, col = "orange")
abline(h = median(total_per_day$total_steps), lty = 2, lwd = 3, col = "black")
abline(h = mean(total_per_day$total_steps), lty = 2, lwd = 3, col = "red")
text(x = 0, y = median(total_per_day$total_steps), pos = 3, labels = "median")
text(x = 0, y = mean(total_per_day$total_steps), pos = 1, labels = "mean", col = "red")
```

### 2. Make a histogram of the total number of steps taken each day
Histogram does not contain days where all observations are missing (i.e. there have to be a number of steps for at least one interval for that day, to be included).   
Otherwise, there would be about 10 days with 0 steps.
```{r histogram}
total_per_day <- filter(total_per_day, na < 1)
hist(total_per_day$total_steps, col = "orange", breaks = 20, main = "Total Steps Per Day", xlab = "Steps Per Day")
abline(v = median(total_per_day$total_steps), lty = 3, lwd = 2, col = "black")
legend(legend = "median", "topright", lty = 3, lwd = 2, bty = "n")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r summary}
mean_steps <- mean(total_per_day$total_steps, na.rm = TRUE)
median_steps <- median(total_per_day$total_steps, na.rm = TRUE)
```
Mean and median of the total number of steps taken per day are `r round(mean_steps, 2)` steps and `r median_steps` steps, respectively.

## What is the average daily activity pattern?
### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, average across all days (y-axis)
```{r daily_1, fig.width = 12}
library(dplyr, quietly = TRUE)
daily_patterns <- activity %>% group_by(interval) %>% summarise(average = mean(steps, na.rm = TRUE))
plot(x = 1:nrow(daily_patterns), y = daily_patterns$average, type = "l", col = "red", xaxt = "n", xlab = "Intervals", ylab = "Average for given interval across all days")
axis(1, labels = daily_patterns$interval[seq(1, 288, 12)], at = seq_along(daily_patterns$interval)[seq(1, 288, 12)])
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r daily_2}
max_number_steps <- filter(daily_patterns, average == max(average))
```
Interval **"`r max_number_steps$interval`"** contains on average the maximum number of steps (**`r round(max_number_steps$average, 2)` steps**).

## Imputing missing values
### 1. Calculate and report the total number of missing value in the dataset (i.e. the total number of rows with NAs)
```{r missing}
na_number <- sum(is.na(activity$steps))
na_number
percentage_na <- mean(is.na(activity$steps))
percentage_na
```
Total number of missing values in the dataset amounts to **`r na_number`** (what is **`r round(percentage_na*100, 1)`** % of the total observations).

### 2. Devise a strategy for filling in all of the missing values in the dataset
As the missing values in the dataset is fairly large, we cannot be sure if there is no bias introduced by missing values. Therefore, we impute missing values based on average number of steps in particular 5-miutes interval.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r na_imputed, cache = TRUE}
without_NAs <- numeric(nrow(activity))
for (i in 1:nrow(activity)){
    if (is.na(activity[i, "steps"]) == TRUE){
        without_NAs[i] <- filter(daily_patterns, interval == activity[i, "interval"]) %>% select(average)
    } else {
        without_NAs[i] <- activity[i, "steps"]
    }
}
activity_without_NAs <- mutate(activity, steps_no_NAs = without_NAs)
head(activity_without_NAs)
```
Below code is just to verify if process of imputing missing values correctly preserved original values (lines with no NAs)
```{r test_NA}
check <- filter(activity_without_NAs, !is.na(steps)) %>% mutate(ok = (steps == steps_no_NAs))
mean(check$ok)
```

### 4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day
```{r histogram_no_NA}
total_day_no_NA <- activity_without_NAs %>% mutate(steps_no_NAs = as.numeric(steps_no_NAs)) %>% group_by(date) %>% summarise(total_steps = sum(steps_no_NAs))
hist(total_day_no_NA$total_steps, col = "blue", breaks = 20, main = "Total Steps Per Day", xlab = "Steps Per Day")
abline(v = median(total_per_day$total_steps), lty = 3, lwd = 2, col ="black")
legend(legend = "median", "topright", lty = 3, lwd = 2, bty = "n")
```

```{r summary_no_NA}
summary(total_day_no_NA$total_steps)
```
Imputing missing values, mean of the total number of steps taken per day increased while median decreased, compared to estimates from the first part (ignoring missing values).  
Imputing missing data resulted in increase of total daily number of steps (instead of each NAs we have average that is always >= 0).


## Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend
```{r weekday}
library(lubridate)
is_weekday <- function(date){
    if(wday(date) %in% c(1, 7)) result <- "weekend"
    else
          result <- "weekday"
    result
}

activity_without_NAs <- mutate(activity_without_NAs, date=ymd(date)) %>% mutate(day = sapply(date, is_weekday))
table(activity_without_NAs$day)
```

### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays or weekends (y-axis)
```{r comparison, cache = TRUE, fig.width = 10}
library(ggplot2)
daily_patterns <- activity_without_NAs %>% mutate(day = factor(day, levels = c("weekend", "weekday")), steps_no_NAs = as.numeric(steps_no_NAs)) %>% group_by(interval, day) %>% summarise(average = mean(steps_no_NAs))
qplot(interval, average, data = daily_patterns, geom = "line", facets = day~.)
```